---
output: github_document
bibliography: "inst/references.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# medRCT

<!-- badges: start -->
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/T0ngChen/medRCT/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/T0ngChen/medRCT/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/T0ngChen/medRCT/graph/badge.svg)](https://app.codecov.io/gh/T0ngChen/medRCT)
<!-- badges: end -->

> Causal **Med**iation Analysis for Estimating Interventional Effects that Emulate a **R**andomised **C**ontrolled **T**rial

The R package `medRCT` supports the estimation of interventional effects [@vanderweele2014effect] that are defined explicitly in terms of a target trial [@hernan2016using], where the treatment strategies are specified to reflect the hypothetical interventions, which are encoded by shifts in the joint distribution of the mediators [@Moreno2021Mediation]. `medRCT` can accommodates multiple mediators in the presence of exposure-induced mediator-outcome confounders.  

## Installation

The `medRCT` package is not yet available on CRAN. You can install the latest stable version from [GitHub](https://github.com/T0ngChen/medRCT) using the following command:

``` r
remotes::install_github("T0ngChen/medRCT")
```

## Example

We here illustrate how to use `medRCT` to estimate interventional indirect effects that emulate a target trial. Specifically, we consider a hypothetical intervention that shifts the distribution of each mediator individually. It allows for the investigation of causal pathways and the quantification of indirect effects to directly address real-world research questions. Consider the following example using simulated data based on the Longitudinal Study of Australian Children:


```{r eg, warning=FALSE, message=FALSE}
# Load the medRCT package
library(medRCT)

# Set a seed for reproducibility
set.seed(2024)

# Display the first few rows of the dataset
head(LSACdata)

# Define confounders for the analysis
confounders <- c("child_sex", "child_atsi", "mat_cob", "mat_engl", "mat_age")

# Estimate interventional indirect effects for a hypothetical intervention
# that shifts the distribution of each mediator individually
med_res <- medRCT(
  dat = LSACdata,                      
  exposure = "sep",                    
  outcome = "child_mh",                
  mediators = c("parent_mh", "preschool_att"), 
  intermediate_confs = "fam_stress",  # intermediate confounders 
  confounders = confounders,           
  bootstrap = TRUE,                    
  intervention_type = "shift_k",       
  mcsim = 100                          
)

# Summarise the results
summary(med_res)
```

For detailed guidance on using the package to handle more complex scenarios, please refer to the [vignette](https://t0ngchen.github.io/medRCT/articles/intro.html).

## Citation

For work involving the `medRCT` R package, please cite the following:

    @software{Chen2024medRCT,
       author = {Tong Chen and Margarita Moreno-Betancur and Ghazaleh Dashti},
       title = {`r gsub("\n", " ", paste0("medRCT: ", utils::packageDescription("medRCT", fields = "Title")))`},
       year  = {`r format(Sys.Date(), "%Y")`},
       url = {https://t0ngchen.github.io/medRCT/},
       note = {R package version `r utils::packageDescription("medRCT", fields = "Version")`}
       }
    @article{Moreno2021Mediation,
       author={Margarita Moreno-Betancur and Paul Moran and Denise Becker and George C Patton and John B Carlin},
       title={Mediation effects that emulate a target randomised trial: Simulation-based evaluation of ill-defined interventions on multiple mediators},
       journal={Statistical Methods in Medical Research},
       volume={30},
       number={6},
       pages={1395--1412},
       year={2021},
       URL={https://doi.org/10.1177/0962280221998409},
       doi={10.1177/0962280221998409},
       publisher={SAGE Publications Ltd}
       }    


## References
