---
title: "Causal Mediation Analysis for Estimating Interventional Effects Mapped to a Target Trial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Mediation Analysis for Estimating Interventional Effects Mapped to a Target Trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "../inst/references.bib"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

Causal mediation analysis methods are commonly used in research studies to explore the causal pathways linking exposures to outcomes. Traditional mediation analysis aims to decompose the total effect of an exposure on an outcome into the natural direct and indirect effects, with the natural indirect effect capturing the exposure effect acting through the mediator. However, the goal of research studies, especially in health and medical research, is often to elucidate the extent to which hypothetical interventions targeting mediators could counter exposure effects. This suggests that a shift in focus is warranted, from effect decomposition to explicitly evaluating hypothetical mediator interventions that may inform future policy, practice and intervention development.

Interventional effects [@vanderweele2014effect] are mediation estimands that are more aligned with this interventional focus while addressing key identifiability challenges associated with natural effects [@Robins1992]. Indeed, interventional effects have been shown to implicitly emulate effects in target trials [@hernan2016using] assessing the impact of distributional shifts in mediators [@MorenoBetancur2018]. Based on this concept, @Moreno2021Mediation proposed definitions of interventional effects that map explicitly to a target trial assessing hypothetical mediator interventions of specific interest, making them particularly useful for policy-relevant evaluations in applied research as illustrated in several published examples [@Dashti2022; @Goldfeld2023; @Afshar2024].

## Interventional effects: definition and estimation methods

@Moreno2021Mediation provided definitions for interventional effects based on three types of hypothetical mediator interventions. In this vignette, we describe these estimands and their corresponding mediator interventions, as implemented in the `medRCT` R package. The type of mediator intervention can be specified using the argument `intervention_type`. The default, `intervention_type = "all"`, estimates all three types of interventional effects, which are:
 
 - **`shift_all`**:  
  This estimand corresponds to an intervention that shifts the **joint distribution of all mediators** in the exposed, given confounders, to match the corresponding distribution in the unexposed.
 - **`shift_k`**:  
  This estimand corresponds to an intervention that shifts the **distribution of a specific mediator** in the exposed, given confounders, to match the corresponding distribution in the unexposed, independent of and **without considering flow-on effects** on other mediators.

 - **`shift_k_order`**:  
  This estimand corresponds to an intervention that shifts the **distribution of a specific mediator** in the exposed, given confounders, to match the corresponding distribution in the unexposed, **while considering flow-on effects** on causally descendant mediators.

`medRCT` estimates interventional effects based on these interventions using a Monte Carlo simulation-based g-computation approach, as described by @Vansteelandt2017. The number of Monte Carlo simulations can be specified using the argument `mcsim`, allowing users to balance computational efficiency and estimation accuracy. The default number of Monte Carlo simulations used is 200. Additionally, `medRCT` automatically removes records with missing data for any of the analysis variables, performing a complete-case analysis.


## Getting started with `medRCT`: Estimating interventional effects

The `medRCT` package is designed to handle complex mediation settings involving **multiple interrelated mediators** and **intermediate confounders** (exposure-induced mediator-outcome confounders). It supports a range of variable types. Specifically, it supports **categorical exposures** with two or more levels (Levels $\geq$ 2). For the **outcome variable**, `medRCT` supports both **binary** and **continuous** outcomes. Additionally, it supports any number of mediators. Each **mediator**, including those not of primary interest (i.e., **intermediate confounders**), can be either **binary** or **continuous**. Any number and type of **baseline confounders** are supported. This makes the package suitable for a broad range of real-world applications in epidemiology, public health, and social science research. 

We begin by illustrating how to estimate the three different types of interventional effects in the presence of multiple mediators and intermediate confounders. Such intermediate confounders can be specified using the argument `intermediate_confs`. We consider the following example using simulated data inspired by a case study within the Longitudinal Study of Australian Children (LSAC) [@Goldfeld2023]. Specifically, we aim to estimate the difference in expected outcome (risk of child mental health problems) under exposure (low socioeconomic position) with versus without a hypothetical intervention that individually shifts the distribution of each mediator (parental mental health and preschool attendance) to the levels in the unexposed (high socioeconomic position), while accounting for baseline confounders, an intermediate (exposure-induced) mediator-outcome confounder and correlations amongst mediators.


```{r, eval=FALSE, echo=FALSE}
library(medRCT)

# Set a seed for reproducibility
set.seed(2025)
# Display the first few rows of the dataset
head(LSACdata)

# Define confounders for the analysis
confounders <- c("child_sex", "child_atsi", "mat_cob", "mat_engl", "mat_age")

# Define intermediate confounders
intermediate_confs <- "fam_stress"

# Estimate interventional indirect effects
med_res <- medRCT(
  dat = LSACdata,                      
  exposure = "sep",                    
  outcome = "child_mh",                
  mediators = c("parent_mh", "preschool_att"), 
  intermediate_confs = intermediate_confs, 
  confounders = confounders,
  interactions_XC = "all",
  intervention_type = "all",
  bootstrap = TRUE,
  boot_args = list(R = 100, stype = "i", ci.type = "norm"),
  mcsim = 200                          
)

# Summarise the results
summary(med_res)
```

<pre class="sourceCode r"><code class="sourceCode r"><span id="cb1-1"><a href="#cb1-1" tabindex="-1"></a><span class="fu">library</span>(medRCT)</span>
<span id="cb1-2"><a href="#cb1-2" tabindex="-1"></a><span class="co">#&gt; medRCT: Causal mediation analysis estimating interventional effects mapped to a target trial</span></span>
<span id="cb1-3"><a href="#cb1-3" tabindex="-1"></a><span class="co">#&gt; Note: When setting intervention_type = 'shift_k_order', the order of the mediators as specified </span></span>
<span id="cb1-4"><a href="#cb1-4" tabindex="-1"></a><span class="co">#&gt; in the 'mediators' argument is important.</span></span>
<span id="cb1-5"><a href="#cb1-5" tabindex="-1"></a></span>
<span id="cb1-6"><a href="#cb1-6" tabindex="-1"></a><span class="co"># Set a seed for reproducibility</span></span>
<span id="cb1-7"><a href="#cb1-7" tabindex="-1"></a><span class="fu">set.seed</span>(<span class="dv">2025</span>)</span>
<span id="cb1-8"><a href="#cb1-8" tabindex="-1"></a><span class="co"># Display the first few rows of the dataset</span></span>
<span id="cb1-9"><a href="#cb1-9" tabindex="-1"></a><span class="fu">head</span>(LSACdata)</span>
<span id="cb1-10"><a href="#cb1-10" tabindex="-1"></a><span class="co">#&gt;   child_sex child_atsi mat_cob mat_engl mat_age sep fam_stress parent_mh</span></span>
<span id="cb1-11"><a href="#cb1-11" tabindex="-1"></a><span class="co">#&gt; 1         0          1       0        0       1   0          0         0</span></span>
<span id="cb1-12"><a href="#cb1-12" tabindex="-1"></a><span class="co">#&gt; 2        NA          0       0        0      NA   0         NA        NA</span></span>
<span id="cb1-13"><a href="#cb1-13" tabindex="-1"></a><span class="co">#&gt; 3        NA          0       0        0      NA   0         NA        NA</span></span>
<span id="cb1-14"><a href="#cb1-14" tabindex="-1"></a><span class="co">#&gt; 4        NA          0       0        0      NA   0         NA        NA</span></span>
<span id="cb1-15"><a href="#cb1-15" tabindex="-1"></a><span class="co">#&gt; 5         1          0       0        0       1   1          0         0</span></span>
<span id="cb1-16"><a href="#cb1-16" tabindex="-1"></a><span class="co">#&gt; 6         1          0       0        0       1   0          1         1</span></span>
<span id="cb1-17"><a href="#cb1-17" tabindex="-1"></a><span class="co">#&gt;   preschool_att child_mh child_SDQscore</span></span>
<span id="cb1-18"><a href="#cb1-18" tabindex="-1"></a><span class="co">#&gt; 1             1        0       8.924660</span></span>
<span id="cb1-19"><a href="#cb1-19" tabindex="-1"></a><span class="co">#&gt; 2             0        0       7.349826</span></span>
<span id="cb1-20"><a href="#cb1-20" tabindex="-1"></a><span class="co">#&gt; 3             0        1      12.824643</span></span>
<span id="cb1-21"><a href="#cb1-21" tabindex="-1"></a><span class="co">#&gt; 4             0        0       6.611369</span></span>
<span id="cb1-22"><a href="#cb1-22" tabindex="-1"></a><span class="co">#&gt; 5             0        1      10.329341</span></span>
<span id="cb1-23"><a href="#cb1-23" tabindex="-1"></a><span class="co">#&gt; 6             0        1      13.552515</span></span>
<span id="cb1-24"><a href="#cb1-24" tabindex="-1"></a></span>
<span id="cb1-25"><a href="#cb1-25" tabindex="-1"></a><span class="co"># Define confounders for the analysis</span></span>
<span id="cb1-26"><a href="#cb1-26" tabindex="-1"></a>confounders <span class="ot">&lt;-</span> <span class="fu">c</span>(<span class="st">"child_sex"</span>, <span class="st">"child_atsi"</span>, <span class="st">"mat_cob"</span>, <span class="st">"mat_engl"</span>, <span class="st">"mat_age"</span>)</span>
<span id="cb1-27"><a href="#cb1-27" tabindex="-1"></a></span>
<span id="cb1-28"><a href="#cb1-28" tabindex="-1"></a><span class="co"># Define intermediate confounders</span></span>
<span id="cb1-29"><a href="#cb1-29" tabindex="-1"></a>intermediate_confs <span class="ot">&lt;-</span> <span class="st">"fam_stress"</span></span>
<span id="cb1-30"><a href="#cb1-30" tabindex="-1"></a></span>
<span id="cb1-31"><a href="#cb1-31" tabindex="-1"></a><span class="co"># Estimate interventional indirect effects</span></span>
<span id="cb1-32"><a href="#cb1-32" tabindex="-1"></a>med_res <span class="ot">&lt;-</span> <span class="fu">medRCT</span>(</span>
<span id="cb1-33"><a href="#cb1-33" tabindex="-1"></a>  <span class="at">dat =</span> LSACdata,                      </span>
<span id="cb1-34"><a href="#cb1-34" tabindex="-1"></a>  <span class="at">exposure =</span> <span class="st">"sep"</span>,                    </span>
<span id="cb1-35"><a href="#cb1-35" tabindex="-1"></a>  <span class="at">outcome =</span> <span class="st">"child_mh"</span>,                </span>
<span id="cb1-36"><a href="#cb1-36" tabindex="-1"></a>  <span class="at">mediators =</span> <span class="fu">c</span>(<span class="st">"parent_mh"</span>, <span class="st">"preschool_att"</span>), </span>
<span id="cb1-37"><a href="#cb1-37" tabindex="-1"></a>  <span class="at">intermediate_confs =</span> intermediate_confs, </span>
<span id="cb1-38"><a href="#cb1-38" tabindex="-1"></a>  <span class="at">confounders =</span> confounders,</span>
<span id="cb1-39"><a href="#cb1-39" tabindex="-1"></a>  <span class="at">interactions_XC =</span> <span class="st">"all"</span>,</span>
<span id="cb1-40"><a href="#cb1-40" tabindex="-1"></a>  <span class="at">intervention_type =</span> <span class="st">"all"</span>,</span>
<span id="cb1-41"><a href="#cb1-41" tabindex="-1"></a>  <span class="at">bootstrap =</span> <span class="cn">TRUE</span>,</span>
<span id="cb1-42"><a href="#cb1-42" tabindex="-1"></a>  <span class="at">boot_args =</span> <span class="fu">list</span>(<span class="at">R =</span> <span class="dv">100</span>, <span class="at">stype =</span> <span class="st">"i"</span>, <span class="at">ci.type =</span> <span class="st">"norm"</span>),</span>
<span id="cb1-43"><a href="#cb1-43" tabindex="-1"></a>  <span class="at">mcsim =</span> <span class="dv">200</span>                          </span>
<span id="cb1-44"><a href="#cb1-44" tabindex="-1"></a>)</span>
<span id="cb1-45"><a href="#cb1-45" tabindex="-1"></a><span class="co">#&gt; Conducting complete case analysis, 2499 observations were excluded due to missing data.</span></span>
<span id="cb1-46"><a href="#cb1-46" tabindex="-1"></a></span>
<span id="cb1-47"><a href="#cb1-47" tabindex="-1"></a><span class="co"># Summarise the results</span></span>
<span id="cb1-48"><a href="#cb1-48" tabindex="-1"></a><span class="fu">summary</span>(med_res)</span>
<span id="cb1-49"><a href="#cb1-49" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-50"><a href="#cb1-50" tabindex="-1"></a><span class="co">#&gt; Estimated interventional indirect effect: </span></span>
<span id="cb1-51"><a href="#cb1-51" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-52"><a href="#cb1-52" tabindex="-1"></a><span class="co">#&gt;                                  Estimate Std. Error  CI Lower  CI Upper</span></span>
<span id="cb1-53"><a href="#cb1-53" tabindex="-1"></a><span class="co">#&gt; IIE_1 (p_trt - p_1)              0.012233   0.003988  0.004156  0.019787</span></span>
<span id="cb1-54"><a href="#cb1-54" tabindex="-1"></a><span class="co">#&gt; IIE_2 (p_trt - p_2)              0.000284   0.002492 -0.004544  0.005224</span></span>
<span id="cb1-55"><a href="#cb1-55" tabindex="-1"></a><span class="co">#&gt; IIE_all (p_trt - p_all)          0.012192   0.004735  0.002680  0.021243</span></span>
<span id="cb1-56"><a href="#cb1-56" tabindex="-1"></a><span class="co">#&gt; IIE_1_prime (p_trt - p_1_prime)  0.013044   0.004151  0.004521  0.020794</span></span>
<span id="cb1-57"><a href="#cb1-57" tabindex="-1"></a><span class="co">#&gt; TCE (p_trt - p_ctr)              0.129132   0.024456  0.082318  0.178184</span></span>
<span id="cb1-58"><a href="#cb1-58" tabindex="-1"></a><span class="co">#&gt;                                 p-value</span></span>
<span id="cb1-59"><a href="#cb1-59" tabindex="-1"></a><span class="co">#&gt; IIE_1 (p_trt - p_1)              0.0022</span></span>
<span id="cb1-60"><a href="#cb1-60" tabindex="-1"></a><span class="co">#&gt; IIE_2 (p_trt - p_2)              0.9092</span></span>
<span id="cb1-61"><a href="#cb1-61" tabindex="-1"></a><span class="co">#&gt; IIE_all (p_trt - p_all)          0.0100</span></span>
<span id="cb1-62"><a href="#cb1-62" tabindex="-1"></a><span class="co">#&gt; IIE_1_prime (p_trt - p_1_prime)  0.0017</span></span>
<span id="cb1-63"><a href="#cb1-63" tabindex="-1"></a><span class="co">#&gt; TCE (p_trt - p_ctr)             1.3e-07</span></span>
<span id="cb1-64"><a href="#cb1-64" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-65"><a href="#cb1-65" tabindex="-1"></a><span class="co">#&gt; Estimated interventional direct effect: </span></span>
<span id="cb1-66"><a href="#cb1-66" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-67"><a href="#cb1-67" tabindex="-1"></a><span class="co">#&gt;                                 Estimate Std. Error CI Lower CI Upper p-value</span></span>
<span id="cb1-68"><a href="#cb1-68" tabindex="-1"></a><span class="co">#&gt; IDE_1 (p_1 - p_ctr)               0.1169     0.0246   0.0700   0.1666 2.1e-06</span></span>
<span id="cb1-69"><a href="#cb1-69" tabindex="-1"></a><span class="co">#&gt; IDE_2 (p_2 - p_ctr)               0.1288     0.0243   0.0823   0.1775 1.1e-07</span></span>
<span id="cb1-70"><a href="#cb1-70" tabindex="-1"></a><span class="co">#&gt; IDE_all (p_all - p_ctr)           0.1169     0.0245   0.0703   0.1663 1.8e-06</span></span>
<span id="cb1-71"><a href="#cb1-71" tabindex="-1"></a><span class="co">#&gt; IDE_1_prime (p_1_prime - p_ctr)   0.1161     0.0247   0.0692   0.1660 2.6e-06</span></span>
<span id="cb1-72"><a href="#cb1-72" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-73"><a href="#cb1-73" tabindex="-1"></a><span class="co">#&gt; Estimated expected outcome in each trial arm:</span></span>
<span id="cb1-74"><a href="#cb1-74" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-75"><a href="#cb1-75" tabindex="-1"></a><span class="co">#&gt;           Estimate Std. Error CI Lower CI Upper p-value</span></span>
<span id="cb1-76"><a href="#cb1-76" tabindex="-1"></a><span class="co">#&gt; p_1         0.3300     0.0225   0.2868   0.3751  &lt;2e-16</span></span>
<span id="cb1-77"><a href="#cb1-77" tabindex="-1"></a><span class="co">#&gt; p_2         0.3419     0.0220   0.2994   0.3858  &lt;2e-16</span></span>
<span id="cb1-78"><a href="#cb1-78" tabindex="-1"></a><span class="co">#&gt; p_all       0.3300     0.0224   0.2871   0.3748  &lt;2e-16</span></span>
<span id="cb1-79"><a href="#cb1-79" tabindex="-1"></a><span class="co">#&gt; p_ctr       0.2131     0.0100   0.1930   0.2323  &lt;2e-16</span></span>
<span id="cb1-80"><a href="#cb1-80" tabindex="-1"></a><span class="co">#&gt; p_trt       0.3422     0.0222   0.2994   0.3864  &lt;2e-16</span></span>
<span id="cb1-81"><a href="#cb1-81" tabindex="-1"></a><span class="co">#&gt; p_1_prime   0.3292     0.0226   0.2860   0.3745  &lt;2e-16</span></span>
<span id="cb1-82"><a href="#cb1-82" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-83"><a href="#cb1-83" tabindex="-1"></a><span class="co">#&gt; Sample Size: 2608 </span></span>
<span id="cb1-84"><a href="#cb1-84" tabindex="-1"></a><span class="co">#&gt; </span></span>
<span id="cb1-85"><a href="#cb1-85" tabindex="-1"></a><span class="co">#&gt; Simulations: 200</span></span></code></pre>

In the summary table:  

- `IIE_all` is the estimated interventional indirect effect when applying the hypothetical intervention specified by `shift_all`. A hypothetical intervention improving both the mental health of parents of children and preschool attendance in low socioeconomic position to the levels in those with high socioeconomic position could potentially prevent 1 per 100 cases of child mental health problems.

- `IIE_k` is the estimated interventional indirect effect when applying the hypothetical intervention described by `shift_k` to a specific mediator `k`. A hypothetical intervention improving the mental health of parents of children (IIE_1) in low socioeconomic position to the levels in those with high socioeconomic position could potentially prevent 1 per 100 cases of child mental health problems. Meanwhile, the effect of a hypothetical intervention on preschool attendance (IIE_2) is negligible.

- `IIE_k_prime` is the estimated interventional indirect effect for a specific mediator `k`, accounting for the flow-on effects on its descendant mediators under the hypothetical intervention described by `shift_k_order`. A hypothetical intervention improving the mental health of parents of children (IIE_1_prime) in low socioeconomic position to the levels in those with high socioeconomic position, while accounting for flow-on effects on preschool attendance, could potentially prevent 1 per 100 cases of child mental health problems. 

- `TCE` is the estimated total causal effect.


## Specifying interactions with `interactions_XC`

The argument `interactions_XC` in `medRCT` allows users to specify two-way interactions amongst exposure and baseline confounders. Of note, the algorithm automatically includes all two-way interactions between the exposure, mediators, and intermediate confounders. These interaction terms are fixed and cannot be modified by the user.

#### Default option:  
- By default, `interactions_XC = "all"` includes **all two-way interactions** between the exposure and baseline confounders, while **baseline confounder-baseline confounder interactions are excluded**.

#### Removing interactions:  
- Setting `interactions_XC = "none"` **removes all two-way interactions** involving the exposure and baseline confounders.  

#### Custom interactions:  
- **Custom specification:** Users can manually specify interaction terms by providing the model formula in the format `interactions_XC = "exposure:confounder1 + exposure:confounder2"`. This option provides full control over interaction terms between the exposure and baseline confounders in the model, enabling customisation for specific research questions.  

## Special cases

#### Analysis without intermediate confounders:

If there are **no intermediate confounders** in the analysis, set the argument **`intermediate_confs = NULL`** when calling the `medRCT` function, as described in the following example:

```{r eval=FALSE}
med_res <- medRCT(
  dat = LSACdata,
  exposure = "sep",
  outcome = "child_mh",
  mediators = c("parent_mh", "preschool_att"),
  intermediate_confs = intermediate_confs,
  confounders = confounders,
  interactions_XC = "all",
  intervention_type = "all",
  intermediate_confs = NULL,
  bootstrap = TRUE,
  boot_args = list(R = 100, stype = "i", ci.type = "norm"),
  mcsim = 200
)
```


#### Single mediator case:  
If there is only **one mediator**, only the `intervention_type` **`shift_k`** can be estimated, as the other options (`shift_all` and `shift_k_order`) require **multiple mediators** to define meaningful interventional effects.

#### Causal ordering for the estimation of the estimand `shift_k_order`:  
When estimating the estimand `shift_k_order`, the **order of mediators** specified in the argument **`mediators`** is important, as it defines the assumed **causal ordering** among the mediators in the analysis. This ordering determines the **causal descendants** for the mediator of interest, providing essential information for **accounting for flow-on effects** after the hypothetical intervention. Specifying an incorrect mediator ordering may result in **biased estimates** for the estimand `shift_k_order`. 

## Model assessment

We have developed a **Shiny** application for convenient model assessment, which can be launched using the command:  

```r
medRCT_shiny(dat = data)
```

Before conducting the mediation analysis, users are encouraged to assess the models fitted by the algorithm using this Shiny application. The Shiny app aims to provide a user-friendly interface to review model summaries and identify potential warnings and errors, ensuring that the models are appropriately specified before proceeding with the analysis. If issues with model fitting are detected, users are encouraged to adjust the exposure-confounder interaction term as needed. However, **mediators or confounders must not be selected based on model fitting results**.

In the Shiny application, after specifying the relevant arguments and clicking the `Run medRCT models` button, the app will fit all regression models required by the `medRCT` algorithm and generate detailed summaries for each model. These summaries support model assessment, allowing users to identify potential issues with fitting what are usually richly-specified models. Additionally, any **warnings** or **error messages** encountered during the model fitting process are highlighted in the app. 


## References
